---
title: "Final Project Report: 2021-2022 European Leagues Player Stats"
author: "Anusha Bhat, Nandini Neralagi, Noelani Phillips, Marcos Pi Marrero"
output:
  pdf_document:
  html_document:
    toc: no
urlcolor: blue
---

***

# Dataset Description
The dataset is sourced from the 2021-2022 European Leagues Player Stats and provides a comprehensive overview of 2,921 male soccer players across prominent leagues, including the Premier League, Ligue 1, Bundesliga, Serie A, and La Liga, during the 2021-2022 season. There are a total of 143 variables. These variables encompass player demographic information, match statistics, passing accuracy, offensive and defensive actions, as well as detailed metrics on aerial duels, tackles, dribbles, and other key aspects of individual and team performance in soccer. The dataset provides comprehensive insights for detailed analysis abd a holistic view of player performances.

The research questions that we will be exploring are:

1. How are player demographics related to leagues and positions?

2. Offense: How does passing performance differ across player position and league, and what are predictors of goals?

3. Defense: How do metrics related to defensive performance differ across player position?


```{r, include = FALSE}
library(countrycode)
library(tidyverse)
library(ggplot2)
futbol = read.csv("~/2021-2022 Football Player Stats.csv", header = TRUE, sep = ";")
extra_match = c("Africa", "Africa", "Europe", "Africa", "Africa", "Americas", "Americas", "Europe", "Africa", "Europe", "Europe", "Africa", "Africa", "Europe", "Europe", "Americas", "Africa", "Americas", "Europe", "Africa", "Europe", "Europe", "Americas", "Asia", "Europe", "Africa", "Europe", "Europe", "Africa", "Americas", "Europe", "Africa", "Africa")
names(extra_match) = c("ALG", "ANG", "BUL", "CGO", "CHA", "CHI", "CRC", "CRO", "CTA", "DEN", "ENG", "EQG", "GAM", "GER", "GRE", "GRN", "GUI", "HON", "KVX", "MAD", "NED", "NIR", "PAR", "PHI", "POR", "RSA", "SCO", "SUI", "TOG", "URU", "WAL", "ZAM", "ZIM")
futbol$Continent = countrycode(futbol$Nation, "iso3c", "continent", custom_match = extra_match)

extra_match2 = c("Algeria", "Angola", "Bulgaria", "Congo", "Chad", "Chile", "Costa Rica", "Croatia", "Central African Republic", "Denmark", "England", "Equatorial Guinea", "Gambia", "Germany", "Greece", "Grenada", "Guinea", "Honduras", "Kosovo", "Madagascar", "Netherlands", "Northern Ireland", "Paraguay", "Philippines", "Portugal", "South Africa", "Scotland", "Switzerland", "Togo", "Uruguay", "Wales", "Zambia", "Zimbabwe")
names(extra_match2) = c("ALG", "ANG", "BUL", "CGO", "CHA", "CHI", "CRC", "CRO", "CTA", "DEN", "ENG", "EQG", "GAM", "GER", "GRE", "GRN", "GUI", "HON", "KVX", "MAD", "NED", "NIR", "PAR", "PHI", "POR", "RSA", "SCO", "SUI", "TOG", "URU", "WAL", "ZAM", "ZIM")

futbol$Country = countrycode(futbol$Nation, "iso3c", "country.name", custom_match = extra_match2)
simpPos = function (x) {
  if (x == "MFFW") {
    return ("MF")
  } else if (x == "FWMF") {
    return ("FW")
  } else if (x == "DFMF") {
    return ("DF")
  } else if (x == "FWDF") {
    return ("FW")
  } else if (x == "MFDF") {
    return ("MF")
  } else if (x == "DFFW") {
    return ("DF")
  } else if (x == "GKMF") {
    return ("GK")
  } else {
    return (x)
  }
}
futbol$Pos_simplified = sapply(futbol$Pos, FUN = simpPos)


futbol$Pos_simplified = sapply(futbol$Pos, FUN = simpPos)
world = map_data("world")

conditions <- c("USA")
replacement_values <- c("United States")
world$region <- replace(world$region, world$region %in% conditions, replacement_values)


futbol$Goals = as.numeric(futbol$Goals)
futbol$Shots = as.numeric(futbol$Shots)


noGoalie = filter(futbol, Pos_simplified=="DF" | Pos_simplified=="FW" | Pos_simplified == "MF")
noGoalieorZero = filter(noGoalie, ShoDist!=0)
```



***


#  How are player demographics related to leagues and positions?



#  How are player demographics related to leagues and positions?

```{r, include=False}
world = filter(world, is.na(subregion))
average_long = world %>% 
  group_by(region) %>% 
  summarise("average long" = mean(long))

average_lat = world %>% 
  group_by(region) %>% 
  summarise("average lat" = mean(lat))
colnames(average_lat) = c("Country", "lat")
colnames(average_long) = c("Country", "long")

futbol = merge(futbol,average_lat, "Country")
futbol = merge(futbol, average_long, "Country")

aggregate = futbol %>% 
  group_by(Comp, Country, long, lat) %>% 
  summarize(NumPlayers=sum(Country==Country))

```
## Graph 1.1a: World Map of Players Nationalities, Colored by League and Sized by Aggregate Number of Players

```{r, warnings=False}
ggplot() +
  geom_map(data=world, map=world, aes(x=long, y=lat, map_id=region), color="grey", fill="white") +
  geom_point(data=aggregate, mapping=aes(x=long, y=lat, color=Comp, size=NumPlayers)) +
  labs(title="Players' Nationalities by League") + xlab("Longitude") + ylab("Latitude") + scale_size(range=c(1,3),
             breaks=c(5,50,100,250),
             labels=c("<10","<50", "<100", ">=100"),
             name = "Number of Players",
             guide="legend") +  scale_color_manual(values = c("Bundesliga" = "red", "La Liga" = "chocolate", "Ligue 1" = "forestgreen", "Premier League" = "deepskyblue", "Serie A" = "purple"))

```
This map of the 2568 players' nationalities visually breaks down which countries are represented in the five leagues in the dataset. The color of a point represents the league the players belong to, and the size of a point represents the aggregate number of players of a given nationality. The main takeaway from this graph is that, while players have diverse nationalities, a large portion of players have some kind of European nationality, as expected. Very little players have nationalities in Oceania or North America, with slightly more in South America and even more in Africa likely due to its proximity to Europe. We can also see that Series A and Premier League are reflected in a multitude of countries, whereas Ligue 1 and La Ligua tend to be comprised of South American and African countries, and Bundesliga seems to come almost exclusively from a few select European Countries.

## Graph 1.1b: Europe Map of Players Nationalities, Colored by League and Sized by Aggregate Number of Players

```{r}
ggplot() +
  geom_map(data=world, map=world, aes(x=long, y=lat, map_id=region), color="grey", fill="white") +
  geom_point(data=aggregate, mapping=aes(x=long, y=lat, color=Comp, size=NumPlayers)) +
  labs(title="Players' Nationalities by League (Europe)") + xlab("Longitude") + ylab("Latitude") + scale_size(range=c(1,7),
             breaks=c(5,10,100,250,400),
             labels=c("0-5","6-10", "10-100", "100>", "251-400"),
             name = "Number of Players",
             guide="legend") +  coord_cartesian(ylim=c(30,72), xlim=c(-10,35)) + scale_color_manual(values = c("Bundesliga" = "red", "La Liga" = "chocolate", "Ligue 1" = "forestgreen", "Premier League" = "deepskyblue", "Serie A" = "purple"))

```

Narrowing down the map to countries in Europe, we can get a better idea about the number of players than our world map. We see that La Liga players tend to primarily have a Spanish nationality, Ligue 1 players tend to primarily have a french Nationality, and Bundesliga players tend to primarily have a German, Austrian, or Hungarian nationality. Further, we see that while the majority of European Serie A players have an Italian nationality, the distribution is spread out across many European countries.

***


#  Offense: How does passing performance differ across player position and league? 

```{r, echo=FALSE}
ggplot(futbol, aes(x = Pos_simplified, y = PasTotCmp., fill = Comp)) +
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Passing Accuracy vs. Player Position, by Soccer Leagues", x = "Player Position", y = "Pass Completion Percentage", fill = "League") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

The box plot displays the distribution of passing completion percentages (`PasTotCmp%`) across different player positions, with the data color-coded by the league to show variations between leagues. The plot shows how passing completion percentages  seems the highest for defenders, whereas forwards seem to have lower passing completion percentages. This might be because of defenders often engage in more controlled passes, while forwards may attempt riskier passes for goal-scoring opportunities.

The boxplots for different positions tend to overlap. Also, there does not seem to be a significant difference between the leagues for the passing accuracy by position, as all boxplots overlap for a given position. This plot seems to be relatively informative as we are easily able to compare the range of differences in passing accuracy by position and league that can help answer the question; however, there are some limitations. There appears to be many outliers, especially with lower passing percentages, for all positions other than goal keeper.


```{r, include=FALSE}
model <- lm(PasTotCmp. ~ Pos_simplified * Comp, data = futbol)
residuals=resid(model)
qqnorm(residuals)         # Q-Q plot for normality
qqline(residuals)         
plot(fitted(model), residuals)   # Residuals vs. Fitted values plot for homogeneity, variances, linearity
```


```{r, echo=FALSE}
model <- aov(PasTotCmp. ~ Pos_simplified * Comp, data = futbol)
summary(model)
```

After checking the Q-Q plot and residuals vs. fitted plot, the data seem to reasonably meet the assumptions for an ANOVA test. The ANOVA results above indicate  significant differences in passing accuracy among different player positions (`Pos_simplified`), as indicated by the F-statistic of 94.623 (p-value < 2e-16), at the standard alpha level of 0.05. The factor `Comp` (representing soccer leagues) also significantly influences passing accuracy, with an F-statistic of 11.827 (p-value 1.57e-09), suggesting variations in passing performance across leagues. Furthermore, there is a significant interaction effect between player position and league (`Pos_simplified:Comp`), denoted by an F-statistic of 2.754 (p-value 0.000999), indicating that the impact of player position on passing accuracy differs across leagues. In summary, the findings suggest that both player position and league affiliation significantly contribute to differences in passing accuracy, and the interaction between these factors should be considered for a comprehensive understanding of the variations.


```{r, echo=FALSE}
scaled_data <- scale(futbol[, c("PasTotCmp", "PasTotDist", "PasTotPrgDist", "PasAss", "PasProg", "PasInt", "PasBlocks")])
unique_positions <- unique(futbol$Pos_simplified)
colors <- setNames(c("red", "blue", "green", "purple"), unique_positions)
pca_result <- prcomp(scaled_data, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x[, 1:2])

shapes <- c(16, 17, 18, 19)
shapes <- setNames(shapes,  unique(futbol$League))

par(mfrow = c(1, 2))
plot(pca_scores, col = adjustcolor(colors[futbol$Pos_simplified],  alpha = 0.4), 
     pch=16,
     main = "PCA Scatter Plot by Position",
     xlab = "Principal Component 1", ylab = "Principal Component 2")

legend("topright", legend =unique_positions, fill = unique(colors))


unique_leagues <- unique(futbol$Comp)
colors <- setNames(c("red", "blue", "green", "purple", "yellow"), unique_leagues)

plot(pca_scores, col = adjustcolor(colors[futbol$Comp],  alpha = 0.4), 
     pch=16,
     main = "PCA Scatter Plot by League",
     xlab = "Principal Component 1", ylab = "Principal Component 2")

legend("topright", legend =unique_leagues, fill = unique(colors))
```

Using 7 passing variables- `PasTotCmp`, `PasTotDist`, `PasTotPrgDist`, `PasAss`, `PasProg`, `PasInt`, and `PasBlocks`- PCA projected the data into the lower-dimensional space defined by the first two principal components. There appears to be some grouping by player positions, suggesting that the principal components may capture patterns related to passing performance across different positions. However, there appears to limited distinction by league in the projection, suggesting that the variability explained may not strongly differentiate passing styles between the considered leagues.


```{r, echo=FALSE}
summary(pca_result)
```

As shown above, the first two principle components explain 65.63% of the variability in the data. This indicates majority of the original information is captured in these two components, suggesting an effective reduction in dimensionality while retaining a considerable amount of the variability present in the seven passing variables. 


# Offense: What are predictors of goals?

# Graph : Density Plot of Average Shot Distance Taken by Players Across Different Positions
```{r}
noGoalieorZero %>% ggplot(aes(x = ShoDist)) +
geom_density(aes(fill = Pos_simplified), alpha = 0.5, adjust = 1) +
  scale_fill_discrete(name = "Position", labels = c("Defender", "Forward", "Midfielder")) + 
  labs(title = "Distribution of Shot Distance across Player Position", x = "Shot Distance (Yards)", y = "Density")
```

The above graph plots the density curves of the average distance players shot from, grouped into their respective positions. At a high level, we can see that players tend to be in the same ballpark in terms of shot distance regardless of player position. There are three main peaks around 10-20 yards. It may seem intuitive to believe that the goal accuracy would increase the closer one gets to the goal, so the fact that there is no concentration of shots right near the goal seems surprising. However, opponent defense may prevent a player from getting as close to the goal as they want or they may strategically choose to shoot from further, so the existence of large peaks further away is not too surprising. What is curious however,is the distribution of defender shot distance. Generally, we expect these players to shoot less often than their counterparts, however it appears that the curve is more spread out and even peaks slightly earlier than forwards. This indicate that they take shots from varying distances, whereas midfielders have high density of shots in the same bandwidth.


***
#  Defense: How do metrics related to defensive performance differ across player position?

```{r, include=False}
futbol$TotalTkl = as.numeric(futbol$TklAtt3rd) + as.numeric(futbol$TklMid3rd) + as.numeric(futbol$TklDef3rd)
futbol$TotalTou = as.numeric(futbol$TouAtt3rd) + as.numeric(futbol$TouMid3rd) + as.numeric(futbol$TouDef3rd)

futbol_nonzero = filter(futbol, TotalTkl>0 & TotalTou>0)
tackle_stats = futbol_nonzero %>% 
  group_by(Pos_simplified) %>% 
  summarize(TklAtt= sum(as.numeric(TklAtt3rd)), TklMid = sum(as.numeric(TklMid3rd)),TklDef= sum(as.numeric(TklDef3rd)), TouAtt= sum(as.numeric(TouAtt3rd)), TouMid= sum(as.numeric(TouMid3rd)), TouDef= sum(as.numeric(TouDef3rd)) )


tackle.long<-melt(tackle_stats)
tackle.long$Zone = c("Attacking 1/3", "Attacking 1/3", "Attacking 1/3", "Attacking 1/3", "Middle 1/3", "Middle 1/3", "Middle 1/3", "Middle 1/3", "Defensive 1/3", "Defensive 1/3", "Defensive 1/3", "Defensive 1/3")
tackle.long$Type = c("Tackles", "Tackles","Tackles","Tackles","Tackles","Tackles","Tackles","Tackles","Tackles","Tackles","Tackles","Tackles", "Touches","Touches","Touches","Touches","Touches","Touches","Touches","Touches","Touches","Touches","Touches","Touches")

```
## Graph : Grouped Bar Graph of Total Touches and Total Tackles Across Player Position and Zone of Field
```{r}
ggplot(tackle.long,aes(Pos_simplified,value,fill=Pos_simplified))+
     geom_bar(stat="identity",position="dodge") + facet_wrap(~Type+Zone, scales="free") + scale_fill_discrete(name = "Position", labels = c("Defender", "Forward", "Goalkeeper", "Midfielder")) + labs(title="Total Player Touches and Tackles by Zone and Position") + xlab("") 
```
This grouped and faceted bar graph examines the number of touches and tackles that players of different positions get in the three zones on a soccer field. Here, a tackle is defined as the taking away of the ball from the feet of an opponent by a player. In this way, the number of tackles is included in the number of touches a player gets. While one purpose of this graph is to examine how touches are distributed across the different positions, it also serves to show us which types of players dominate in certain zones of the field. As expected, we see that the players who are meant to play primarily in a given third of the field tend to get the most touches on the ball. However, the breakdown of tackles is slightly more interesting. Most notably, we notice that midfielders have a comparatively high performance in all areas of the field, whereas the number of tackles that defensive and forward players get tends to decrease the further they get from their assigned zone. Still, while the number of tackles achieved by defensive players in the attacking third is quite close to that of forward players, the number of tackles achieved by forward players in the defensive zone is much smaller than that of defensive players. One additional piece of information gleaned by this graph is that the number of touches in the attacking zone by forwards and midfielders are very similar but the number of tackles by midfielders is higher, indicating that midfielders tend to play a more defensive role in the attacking zone.



***

# Conclusion and Main Takeaways

[add]


***

# Sources

+  [Kaggle](https://www.kaggle.com/datasets/vivovinco/20212022-football-player-stats)
